---
title: "Homework 2"
author: "Harshetha Haritharan - hh29685"
output: pdf_document
date: "2025-01-28"
---
GitHub Repository: https://github.com/harshetha-haritharan/SDS315_HW2

```{r, include=FALSE}
# Import necessary library
library(tidyverse)
```

#  Problem 1: Beauty, or not, in the classroom
### Part A
```{r, echo=FALSE}
# Load data
profs <- read_csv("profs.csv", show_col_types = FALSE)

# Create histogram to show distribution of evaluation scores
ggplot(profs, aes(x = eval)) + geom_histogram(binwidth = 0.25, color = "black",na.rm = TRUE)  +
labs(title = "Distribution of Course Evaluation Scores",x = "Evaluation Score",y = "Frequency")

# Calculate relevant summary statistics
eval_median <- median(profs$eval, na.rm = TRUE)
eval_iqr <- IQR(profs$eval, na.rm = TRUE)
```

The histogram shows the distribution of course evaluation scores that is slightly skewed to the left. The IQR is 0.8 and the median is 4 and these relevant summary statistics indicate that there were typically positive evaluations for the professors. 

### Part B
```{r, echo=FALSE}
# Create boxplots for evaluation scores based on native English speakers or not
ggplot(profs, aes(x = as.factor(native), y = eval, fill = as.factor(native))) + geom_boxplot() +labs(title = "Evaluation Scores by Native English Speaker",x = "Professor's English Proficiency",y = "Evaluation Score",fill = "English Proficiency") +scale_x_discrete(labels = c("No" = "Non-Native Speaker", "Yes" = "Native Speaker")) 

# Calculate relevant summary statistics
mean_native <- mean(profs$eval[profs$native == "yes"], na.rm = TRUE)
sd_native <- sd(profs$eval[profs$native == "yes"], na.rm = TRUE)
mean_nonnative <- mean(profs$eval[profs$native == "no"], na.rm = TRUE)
sd_nonnative <- sd(profs$eval[profs$native == "no"], na.rm = TRUE)
```


The boxplots are not skewed as the median is the middle of each box. The relevant summary statistics for native English speakers are a mean of 4 and a standard deviation of 0.6 and for non-native English speakers the mean is 3.7 and a standard deviation of 0.5. This indicates that native English speakers have a distribution that tends to result in a higher evaluation score than non-native English speakers. 

### Part C

```{r, echo=FALSE}
# Create faceted histogram for evaluation scores by gender
ggplot(profs, aes(x = eval)) + geom_histogram(binwidth = 0.25, color = "black") + facet_wrap(~ gender, nrow = 2) + labs (title = "Distribution of Evaluation Scores by Gender",x = "Evaluation Score",y = "Frequency") 

# Calculate relevant summary statistics
median_male <- median(profs$eval[profs$gender == "male"], na.rm = TRUE)
iqr_male <- IQR(profs$eval[profs$gender == "male"], na.rm = TRUE)

# Calculate median and IQR for female instructors
median_female <- median(profs$eval[profs$gender == "female"], na.rm = TRUE)
iqr_female <- IQR(profs$eval[profs$gender == "female"], na.rm = TRUE)
```

The faceted histograms show a comparision of the distribution between male and female professors’ evaluation scores and both of them have a distribution skewed to the left. The median for males is 4.1 and the IQR is 0.8 and the median for females is a 3.9 and the IQR is a 0.7. This indicates that male professors have a distribution that has scores that are slightly higher than the female professors.

### Part D
```{r, echo=FALSE}
# Create a scatterplot of beauty vs. evaluation scores
ggplot(profs, aes(x = beauty, y = eval)) +geom_point(alpha = 0.6) + labs(title = "Association Between Professor's Physical Attractiveness and Evaluations", x = "Professor's Physical Attractiveness ",y = "Evaluation Score")

# Calculate relevant summary statistics
correlation <- cor(profs$beauty, profs$eval, use = "complete.obs")
```

This scatterplot represents the extent to which there may be an association between the
professor’s physical attractiveness (x) and their course evaluations (y). The calculated correlation coefficient is 0.19 which indicates that there is a weak positive relationship between the two variables presented. 

# Problem 2: Bike Sharing

### Plot A
```{r, echo=FALSE}
# Import data
bike_data <- read_csv("bikeshare.csv",  show_col_types = FALSE)

# Calculate average total bike rentals for each hour of the day
avg_hourly <- bike_data %>%
  group_by(hr) %>%
  summarize(avg_total = mean(total, na.rm = TRUE), .groups = "drop") %>%
  arrange(hr)

# Create line graph showing average hourly bike rentals across all hours
ggplot(avg_hourly, aes(x = hr, y = avg_total)) + geom_line(linewidth = 1) + labs(title = "Average Hourly Bike Rentals",x = "Hour of the Day",y = "Average Rentals") 
```

The line graph shows the average hourly bike rentals (total) across all hours of the day (hr). The graphs indicates that around the times 8 AM and 5 PM, the highest amount of average bike rentals can be found as people are more likely to be moving around their area at these times. Around 4 AM and 11 PM is where the least amount of average bike rentals can be found as people are less likely to be moving around their area at these times. This graph indicates that there are two peak times where average bike rentals are the highest, most likely relating to the time of day people to go work or go back home. 


### Plot B
```{r, echo=FALSE}
# Calculate average bike rentals per hour grouped by working day
avg_hourly_working <- bike_data %>%
  group_by(hr, workingday) %>%
  summarize(avg_total = mean(total, na.rm = TRUE), .groups = "drop") %>%
  mutate(workingday_label = c("Non-Working Day", "Working Day")[workingday + 1]) %>%
  arrange(hr)

# Create the faceted line graph
ggplot(avg_hourly_working, aes(x = hr, y = avg_total)) + geom_line(linewidth = 1) + facet_wrap(~workingday_label, nrow = 2) + labs(title = "Average Bike Rentals by Hour of the Day",x = "Hour of Day",y = "Average Bike Rentals")
```

The faceted line graphs show average hourly bike rentals on working days versus non-working days. On non-working days there is no significant peak as there is a demand throughout the day, especially near the afternoon to evening times as more people enjoy leisurely activities at this time. On work days, the highest average of bikes being rented is around 8 AM and 5 PM while the lowest is around 4 AM. These faceted line graphs indicate that on non-working days there is no significant peak of average bike rentals while working days have two peak times of bike rentals, most likely in relation to the time of day people go to work or back home. 

### Plot C
```{r, echo=FALSE}
# Filter for 9 AM
ridership_9am <- bike_data %>%
  filter(hr == 9, weathersit != 4) %>%  # Remove Heavy Rain, Snow (weathersit = 4)
  group_by(weathersit, workingday) %>%
  summarize(avg_total = mean(total, na.rm = TRUE), .groups = "drop")

# Define weather labels
weather_labels <- c(
  "Clear, Partly Cloudy",
  "Mist, Cloudy",
  "Light Snow, Light Rain"
)

# Create the faceted bar plot
ggplot(ridership_9am, aes(x = factor(weathersit, levels = 1:3, labels = weather_labels), y = avg_total, fill = factor(weathersit))) + geom_bar(stat = "identity") + facet_wrap(~workingday, labeller = as_labeller(c(`0` = "Non-Working Day", `1` = "Working Day"))) + labs(title = "Average Bike Rentals at 9 AM by Weather and Day Type",x = "Weather Situation",y = "Average Bike Rentals",fill = "Weather") + theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

The bar plots are faceted by working day or non-working day and show the average number of bike rentals at 9 AM across different weather situations. The highest frequency of average bike rentals can be seen when there is nicer weather such as clear or cloudy. This pattern can be seen in both working and non-working days, but working days tend to have more average bike rentals for all weather conditions. This indicates that more people are likely to use rental bikes when the weather is nicer outside and on a working day. 

#  Problem 3: Capital Metro UT Ridership
```{r, include=FALSE}
# Load data 
capmetro_UT <- read_csv("capmetro_UT.csv")

# Recode the categorical variables in sensible, rather than alphabetical, order
capmetro_UT <- mutate(capmetro_UT,
  day_of_week = factor(day_of_week, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
  month = factor(month, levels = c("Sep", "Oct", "Nov"))
)
```


### Plot 1

```{r, echo=FALSE}
# Calculate average boardings by hour, day of week, and month
avg_boardings <- capmetro_UT %>%
  group_by(hour_of_day, day_of_week, month) %>%
  summarize(avg_boarding = mean(boarding, na.rm = TRUE), .groups = "drop")

# Create faceted line graph to show average boardings
ggplot(avg_boardings, aes(x = hour_of_day, y = avg_boarding, color = month)) + geom_line(linewidth = 1) + facet_wrap(~day_of_week, scales = "free_x") + labs(title = "Average Boardings by Hour, Day of Week, and Month", x = "Hour of the Day", y = "Average Boardings",color = "Month") 
```


The faceted line graph shows the average number of boardings by hour of the day during September, October, and November, faceted by each day of the week. The hour of peak boardings is broadly similar across the weekdays as they happen around 5 PM, most likely because those are the times people commute the most, and this is seen by the peaks in the graphs. Average boardings on Mondays in September are lower compared to other months and days most likely because of the semester just starting or even Labor Day holiday and this is seen on the Monday graph as the red line is lower than the blue and green line. Average boardings during Wednesdays, Thursdays, and Fridays in November are lower most likely because of the Thanksgiving holiday as many people go back home and this is seen in the graph as all the weekday graphs usually have the blue line below the green and red ones. 


### Plot 2
```{r, echo=FALSE}
# Create column to label weekends and weekdays
capmetro_UT <- mutate(capmetro_UT,
  weekend_label = ifelse(weekend == "weekend", "Weekend", "Weekday")
)

# Create faceted scatter plot of boardings vs temperature
ggplot(capmetro_UT, aes(x = temperature, y = boarding, color = weekend_label)) + geom_point(alpha = 0.6) + facet_wrap(~hour_of_day) + labs(title = "Boardings vs. Temperature by Hour of Day",x = "Temperature (°F)",y = "Boardings",color = "Day Type") 
```


The faceted line graph shows the average number of boardings on weekdays and weekends compared to the temperature is degrees Fahrenheit, faceted by hour of the day. It can be seen that there are a significantly higher number of red dots, indicating that there are more boarding done on weekdays, regardless of the temperature and hour of the day. When we hold hour of day and weekend status constant, temperature does have a noticeable effect on the number of UT students riding the bus because the number of weekend boarding (blue dots) are higher during the warmer temperatures, such as afternoon times. 


\vspace{6cm}  

# Problem 4: Wrangling the Billboard Top 100

### Part A
```{r, echo=FALSE}
# Load data
billboard <- read_csv("billboard.csv", show_col_types = FALSE, col_select = -1, name_repair = "minimal")

# Calculate total weeks for each song
top_songs <- billboard %>%
  group_by(performer, song) %>%
  summarize(count = n(), .groups = "drop") %>%
  arrange(desc(count)) %>%
  slice_head(n = 10)

# Display the table
knitr::kable(top_songs, col.names = c("Performer", "Song", "Count (Week on Chart)"))
```


The table shows the top 10 most popular songs since 1958 and the popularity is measured by the total number of weeks each song spent on the Billboard Top 100. The table includes the performer, song title, and the total number of weeks the song spent on the chart. 

### Part B

```{r, echo=FALSE}
# Count unique songs per year 
musical_diversity <- billboard %>%
  filter(year != 1958 & year != 2021) %>%  
  group_by(year) %>%
  summarize(unique_songs = length(unique(song)), .groups = "drop")  

# Create  line graph
ggplot(musical_diversity, aes(x = year, y = unique_songs)) + geom_line(linewidth = 1) + labs(
title = "Musical Diversity of the Billboard Top 100 Over Time", x = "Year", y = "Number of Unique Songs") 
```

The line graph shows the number of unique songs that appeared on the Billboard Top 100 each year, excluding 1958 and 2021. It can be seen that musical diversity had a significant decline but is now increasing over time with some peaks around the late 1960s and early 2020s. These changes in musical diversity can be caused by trending figures at the time and what type of music is popular. An observation can be made is that during the start of Covid, 2020, there was a peak in musical diversity as more artists had more time at home to spend on their musical craft to create more songs.


### Part C

```{r, echo=FALSE}
# Identify ten-week hits and artists with at least 30 such hits
ten_week_hits <- billboard %>%
  group_by(performer, song) %>%
  summarize(total_weeks = n(), .groups = "drop") %>%
  filter(total_weeks >= 10) %>%
  group_by(performer) %>%
  summarize(ten_week_hits = n(), .groups = "drop") %>%
  filter(ten_week_hits >= 30) %>%
  arrange(desc(ten_week_hits))

# Create bar plot of ten-week hits by artist
ggplot(ten_week_hits, aes(x = fct_reorder(performer, ten_week_hits), y = ten_week_hits, fill = performer)) + geom_bar(stat = "identity", show.legend = FALSE) + coord_flip() + labs(title = "Artists with At Least 30 Ten-Week Hits", x = "Artist", y = "Number of Ten-Week Hits") 
```

The bar plot shows the number of ten-week hits, which are songs that stayed on the Billboard Top 100 for at least 10 weeks, for artists with at least 30 such hits. This chart includes some of the most famous names in music including Elton John, who has the highest number of ten-week hits with over 50 of them and other popular artists as well including Drake and Chicago. 












